---
title: "An Introduction to lambertReg"
author: "Bahadir Yuzbasi"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    css: vignette.css
vignette: >
  %\VignetteIndexEntry{An Introduction to lambertReg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
  fig.width = 9, fig.height = 5.8, out.width = "100%", dpi = 120)
```

# Introduction

**lambertReg** fits sparse Gaussian regression with the Lambert penalty.
The main functions are `lambert()` for a regularization path and
`cv.lambert()` for choosing the penalty strength by cross-validation.
Both provide coefficient extraction, prediction, and plotting methods.
This guide follows one example from fitting through prediction on new data.

For a training sample with \(n\) observations and \(p\) predictors, the
package targets the penalized least-squares objective
\[
  Q_\lambda(b)=\frac{1}{2n}\lVert y_c-Zb\rVert_2^2+
  \sum_{j=1}^p p_\lambda(|b_j|).
\]
Here \(b\in\mathbb R^p\) contains standardized coefficients, \(y_c\) is the
centered response, and \(Z\) is the centered and scaled training matrix.
The intercept is unpenalized. The parameter \(\lambda>0\) controls the
strength of the Lambert penalty \(p_\lambda\), whose shape is fixed at
\(c=1\). The penalty becomes constant at coefficient magnitudes of
\(e\lambda\) and above. Cyclic coordinate updates seek a stationary
solution; a global minimum is not guaranteed for this nonconvex objective.

# Installation

The package requires R (>= 4.1.0), Rcpp, and an R-compatible C++ compiler.
Install from a local clone of the
[GitHub repository](https://github.com/byuzbasi/lambertReg):

```{bash, eval=FALSE}
git clone https://github.com/byuzbasi/lambertReg.git
R CMD INSTALL lambertReg
```

To build this guide as an installed vignette, use `R CMD build lambertReg`
and install the resulting source archive. Vignette building additionally
requires knitr, rmarkdown, and Pandoc.

# Quick start

## Prepare the data

The example has 100 observations, eight predictors, and three nonzero
coefficients. We use 80 observations for training and reserve 20 for
prediction. The test observations do not enter preprocessing or tuning.

```{r data}
library(lambertReg)
set.seed(42)
x <- matrix(rnorm(100 * 8), nrow = 100, ncol = 8)
colnames(x) <- paste0("x", seq_len(ncol(x)))
y <- 2 * x[, 1] - x[, 2] + 0.5 * x[, 3] + rnorm(100)
train <- seq_len(80)
x_train <- x[train, , drop = FALSE]
y_train <- y[train]
x_test <- x[-train, , drop = FALSE]
y_test <- y[-train]
```

## Fit a regularization path

Supply a numeric predictor matrix and a response vector to `lambert()`.
The default path has 60 logarithmically spaced penalty fractions from 1
to 0.001, relative to the training entry score \(\lambda_{\max}\).

```{r path-fit}
path <- lambert(x_train, y_train)
print(path)
```

`plot()` displays the coefficient paths on the original predictor scale.
The intercept is omitted. The horizontal axis decreases from left to
right, following the path from stronger to weaker regularization.

```{r path-plot, fig.alt="Original-scale Lambert coefficient paths as regularization decreases."}
plot(path, labels = TRUE)
```

A path fit does not automatically choose a penalty. Use `index` to extract
coefficients or predictions at a stored path position:

```{r path-extraction}
knitr::kable(coef(path, index = 20L), col.names = "Coefficient", digits = 4)
predict(path, newx = x_test[1:3, , drop = FALSE], index = 20L)
```

## Choose the penalty by cross-validation

`cv.lambert()` uses five folds by default. Supplying `foldid` makes the
partition explicit and reusable. Each fit uses four folds for training
and the remaining fold for validation.

```{r cv-fit}
foldid <- sample(rep(1:5, length.out = length(y_train)))
fit <- cv.lambert(x_train, y_train, foldid = foldid)
print(fit)
stopifnot(fit$ok)
```

The selected candidate minimizes the arithmetic mean of the fold mean
squared errors (MSEs). `lambda.min` is the selected penalty for the full
training data; it does not mean the smallest penalty in the grid.
`index` identifies its path position and `fraction.min` its relative scale.

```{r selected-fit}
knitr::kable(data.frame(
  Lambda = fit$lambda.min,
  Fraction = fit$fraction.min,
  CV_MSE = fit$cvm[fit$index],
  Selected = fit$nzero[fit$index]
), digits = 4, row.names = FALSE)
```

## Plot the CV error

```{r cv-plot, fig.alt="Lambert CV error, fold-based error bars, variable counts, and the selected penalty."}
plot(fit)
```

The horizontal axis is \(\log(\lambda/\lambda_{\max})\), increasing from
weaker to stronger regularization. Blue points show the mean fold MSE;
the orange point and dashed line mark the selected penalty. The upper
axis counts selected variables in each full-training model, excluding
the intercept. A coefficient is counted when its original-scale magnitude
exceeds \(10^{-8}\).

Error bars show one descriptive standard error: the standard deviation
of the \(K\) fold MSEs divided by \(\sqrt K\), where \(K\) is the number
of folds. Because training samples overlap, these are not
independent-sample confidence intervals. Selection uses the minimum CV
error, rather than a one-standard-error rule.

For a compact figure, use `style = "paper"`. Other display options change
the same stored results without refitting:

```{r plot-options, eval=FALSE}
plot(fit, style = "paper")
plot(fit, xaxis = "lambda")    # log full-training lambda
plot(fit, error_bars = FALSE)
plot(fit, show_nzero = FALSE)
```

## Extract the selected model

`coef(fit)` returns the intercept and selected coefficients in the original
units. No post-selection least-squares refit is applied.

```{r coefficients}
knitr::kable(coef(fit), col.names = "Coefficient", digits = 4)
```

The coefficient plot from a CV fit also marks the selected penalty:

```{r coefficient-plot, fig.alt="Lambert coefficient paths with a vertical marker at the CV-selected penalty."}
plot(fit, type = "coefficients", labels = TRUE, xaxis = "lambda")
```

## Predict on new observations

Supply new predictors in the same column order and with the same names as
in training. Pass their original values; the fitted object incorporates
the training transformation and predicts on the original response scale.

```{r prediction}
pred <- drop(predict(fit, newx = x_test))
knitr::kable(head(data.frame(Observed = y_test, Predicted = pred)), digits = 3)
knitr::kable(data.frame(
  RMSE = sqrt(mean((y_test - pred)^2)),
  MAE = mean(abs(y_test - pred))
), digits = 4, row.names = FALSE)
```

These held-out errors illustrate prediction on new observations. For a
method comparison, use an independent test sample or outer
cross-validation; the tuning-CV minimum is not an external test error.

# Penalty scale and preprocessing

Within each training sample, predictor \(j\) is centered by its mean
\(\bar x_j\) and divided by its centered root-mean-square value
\(s_j=\{n^{-1}\sum_{i=1}^n(x_{ij}-\bar x_j)^2\}^{1/2}\).
Thus \(Z_{ij}=(x_{ij}-\bar x_j)/s_j\) has squared column norm \(n\).
The response is centered by its training mean and is not variance-scaled.
CV estimates these quantities using only the training observations of
each fold.

The entry score \(\lambda_{\max}=\max_j|Z_j^\top y_c|/n\), where \(Z_j\)
is column \(j\) of the standardized design, defines the upper end of the
relative path. For candidate fraction \(\tau_k\in(0,1]\) in fold \(f\),
the penalty is \(\lambda_{k,f}=\tau_k\lambda_{\max,f}\).
CV therefore compares common fractions, while absolute penalties can
differ between folds. The score averages fold MSEs arithmetically, including
when folds have unequal sizes. Numerical ties favor the larger fraction.

A custom fractional grid must be positive, strictly decreasing, and no
larger than one. Absolute penalties can instead be supplied to `lambert()`:

```{r custom-grid, eval=FALSE}
fractions <- 10^seq(0, -3, length.out = 80L)
fit_custom <- cv.lambert(x_train, y_train, foldid = foldid,
                        lambda_fraction = fractions)
path_absolute <- lambert(x_train, y_train, lambda = path$lambda)
```

The scalar penalty and coordinate threshold are also available directly.
For magnitude \(t\geq0\), `lambert_penalty(t, lambda)` evaluates
\(p_\lambda(t)\); `lambert_threshold(z, lambda)` applies the scalar update
to the partial-residual score \(z\) for a standardized coordinate of
unit curvature.

```{r penalty-tools}
z <- c(-3, -1, 0, 1, 3)
knitr::kable(data.frame(
  Score = z,
  Penalty_at_magnitude = lambert_penalty(abs(z), lambda = 1),
  Updated_coefficient = lambert_threshold(z, lambda = 1)
), digits = 4, row.names = FALSE)
```

# Diagnostics and reproducibility

```{r diagnostics}
knitr::kable(data.frame(
  Selected_fit_ok = fit$ok,
  Status = fit$status,
  Partial_CV_search = fit$partial_search,
  Full_path_complete = fit$full_path_complete
), row.names = FALSE)
```

`ok` describes the selected fit, `partial_search` records unavailable CV
candidates, and `full_path_complete` describes the full-training path.
For details, inspect `cv_diagnostics` or `lambert.fit$diagnostics`.
Unverified fits cannot be used by `predict()` or `coef()` and appear as
gaps in path plots.

Each penalty position considers warm and zero starts, retaining the
verified candidate with the lower objective. Exactly identical initial
vectors can share a verified solve; the `executed` and `reused` columns
in `attempts` record this. `max_sweeps` and `kkt_tol` control the iteration
budget and normalized stationarity tolerance. The default values are
10,000 sweeps and \(10^{-7}\); active-curvature diagnostics provide an
additional check. The help pages describe the acceptance criteria.

Inputs must be finite dense numeric matrices with columns that vary in
every training sample. The current interface supports Gaussian responses;
it does not support observation weights or automatic missing-value handling.
Save the folds, input column order, package version, and controls with an
analysis. Supplied folds do not consume random numbers; otherwise folds
are sampled from the caller's random-number state. Early stopping is not
applied.

# Further information

```{r help, eval=FALSE}
?lambert
?cv.lambert
?lambert_penalty
citation("lambertReg")
```

Source code and issue reporting are available at
[byuzbasi/lambertReg](https://github.com/byuzbasi/lambertReg).
The package is distributed under GPL-3.
