---
title: "An Introduction to lambertReg"
author: "Bahadir Yuzbasi"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{An Introduction to lambertReg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
  fig.width = 9, fig.height = 5.8, out.width = "100%", dpi = 120)
```

## Introduction

**lambertReg** fits sparse Gaussian regression with the Lambert penalty.
It provides a regularization path, cross-validation, predictions on the
original response scale, and plots of CV error and coefficient paths.
The penalty shape is fixed at \(c=1\); only the penalty strength is tuned.

For a training sample of size \(n\) with \(p\) predictors, the fitted
standardized coefficient vector \(b\in\mathbb R^p\) minimizes the objective
\[
  \frac{1}{2n}\lVert y_c-Zb\rVert_2^2+
  \sum_{j=1}^p p_\lambda(|b_j|).
\]
Here \(y_c\) is the centered response, \(Z\) is the centered and scaled
training matrix, and \(\lambda>0\) controls regularization. The intercept
is unpenalized. The bounded penalty \(p_\lambda\) shrinks small coefficients
and becomes constant for coefficient magnitudes at least \(e\lambda\).
Its analytic coordinate update avoids a numerical scalar optimization at
each step. The objective is nonconvex: the numerical checks certify
stationarity and assess active curvature, not global optimality.

## Quick start

This small synthetic example has eight predictors and three nonzero
coefficients. We reserve twenty observations for illustrating prediction;
they do not enter preprocessing or tuning.

```{r data}
library(lambertReg)
set.seed(42)
x <- matrix(rnorm(100 * 8), nrow = 100, ncol = 8)
colnames(x) <- paste0("x", seq_len(ncol(x)))
y <- 2 * x[, 1] - x[, 2] + 0.5 * x[, 3] + rnorm(100)
train <- seq_len(80)
x_train <- x[train, , drop = FALSE]
y_train <- y[train]
x_test <- x[-train, , drop = FALSE]
y_test <- y[-train]
```

We supply the fold labels explicitly so that the same partition can be
reused. With five folds, each fit uses four folds for training and one for
validation. The default path contains sixty penalty fractions from 1 to
0.001, spaced uniformly on a logarithmic scale.

```{r cv-fit}
foldid <- sample(rep(1:5, length.out = length(y_train)))
fit <- cv.lambert(x_train, y_train, foldid = foldid)
print(fit)
stopifnot(fit$ok)
```

`lambda.min` is the **CV-selected full-training penalty**, not the smallest
penalty in the grid. `fit$index` identifies its position. After tuning, the
package fits the full-training path, which also supplies the variable counts
and coefficient paths shown below.

```{r selected-fit}
knitr::kable(data.frame(
  Lambda = fit$lambda.min,
  Fraction = fit$fraction.min,
  CV_MSE = fit$cvm[fit$index],
  Selected = fit$nzero[fit$index]
), digits = 4, row.names = FALSE)
```

## Cross-validation plot

```{r cv-plot, fig.alt="Lambert CV error with fold-based error bars, selected-variable counts and the selected penalty."}
plot(fit)
```

The horizontal axis is \(\log(\lambda/\lambda_{\max})\), increasing from
weaker to stronger regularization. Blue points show mean validation-fold
MSE. The orange point and dashed line mark the selected candidate. The upper
axis gives the number of selected variables in each full-training model,
excluding the intercept; selection uses \(|\widehat\beta_j|>10^{-8}\) on
the original coefficient scale. These counts are not fold averages.

Error bars are the mean plus or minus one descriptive standard error,
computed as the standard deviation of the \(K\) fold MSEs divided by
\(\sqrt K\), where \(K\) is the number of folds. Training samples overlap,
so the bars should not be read as independent-sample confidence intervals.
They do not change the minimum-CV selection rule.

For a publication figure, `style = "paper"` omits the title and summary
annotations. The following calls customize the same stored results; none
refits the model.

```{r plot-options, eval=FALSE}
plot(fit, style = "paper")
plot(fit, xaxis = "lambda")    # log full-training lambda
plot(fit, error_bars = FALSE)
plot(fit, show_nzero = FALSE)
```

## Coefficients and regularization paths

`coef()` returns the intercept and coefficients on the original predictor
scale. There is no post-selection least-squares refit.

```{r coefficients}
knitr::kable(coef(fit), col.names = "Coefficient", digits = 4)
```

```{r coefficient-plot, fig.alt="Original-scale Lambert coefficient paths and the CV-selected penalty."}
plot(fit, type = "coefficients", labels = TRUE, xaxis = "lambda")
```

The coefficient plot excludes the intercept and follows the path from
stronger to weaker regularization, so its horizontal axis decreases from
left to right. The dashed line identifies the selected penalty. Curve
magnitudes depend on predictor units; they are not scale-free measures of
importance. Unverified path positions appear as gaps.

If cross-validation is unnecessary, `lambert()` fits a path directly. Use
an explicit index to extract one of its verified fits; coefficients are
not interpolated between path positions.

```{r direct-path}
path <- lambert(x_train, y_train, lambda_fraction = c(1, 0.5, 0.2, 0.1))
print(path)
coef(path, index = 3L)
```

## Prediction on new observations

New predictors must have the same columns, names and order as the training
matrix. Pass their original values: the fitted object handles the training
transformation and returns predictions on the original response scale.

```{r prediction}
pred <- drop(predict(fit, newx = x_test))
knitr::kable(head(data.frame(Observed = y_test, Predicted = pred)), digits = 3)
knitr::kable(data.frame(
  RMSE = sqrt(mean((y_test - pred)^2)),
  MAE = mean(abs(y_test - pred))
), digits = 4, row.names = FALSE)
```

This small held-out example demonstrates the API, not comparative method
performance. For an empirical assessment, use independent test observations
or outer cross-validation. The minimum tuning-CV error is not an external
test-error estimate.

## Preprocessing and the CV scale

Within each training sample, predictor \(j\) is centered by its training
mean \(\bar x_j\) and divided by
\(s_j=\{n^{-1}\sum_i(x_{ij}-\bar x_j)^2\}^{1/2}\).
Thus \(Z_{ij}=(x_{ij}-\bar x_j)/s_j\) has squared column norm \(n\).
The response is centered by its training mean and is not variance-scaled.
These operations are repeated using only the training observations of each
CV fold. Validation observations never determine centers, scales or penalties.

The entry score is \(\lambda_{\max}=\max_j|Z_j^\top y_c|/n\), where
\(Z_j\) denotes column \(j\). CV aligns candidates by a common fraction
\(\tau_k\in(0,1]\). In training fold \(f\), candidate \(k\) uses
\(\lambda_{k,f}=\tau_k\lambda_{\max,f}\); the absolute penalties can
therefore differ between folds. The tuning score is the arithmetic mean of
fold MSEs, even when fold sizes differ. Ties within the documented numerical
tolerance favor the larger fraction. See `?cv.lambert` for the exact rule.

## Checking a fitted object

```{r diagnostics}
knitr::kable(data.frame(
  Selected_fit_ok = fit$ok,
  Status = fit$status,
  Partial_CV_search = fit$partial_search,
  Full_path_complete = fit$full_path_complete
), row.names = FALSE)
```

These fields answer different questions. `ok` concerns the selected fit;
`partial_search` records whether any CV candidate was unavailable; and
`full_path_complete` concerns the full-training path. A partial search can
still yield a verified selected fit. Inspect `fit$cv_diagnostics` for fold
records and `fit$lambert.fit$diagnostics` for full-training records.
`predict()` and `coef()` refuse unverified requested fits.

Each penalty position considers both a warm start and a zero start.
When the initial vectors are exactly identical, a verified warm-start result
is reused; both records remain in `attempts`, with `executed` and `reused`
flags. Distinct starts are solved separately.
The lower-objective verified candidate is retained. `max_sweeps` and
`kkt_tol` control the iteration budget and normalized KKT tolerance; increasing
the budget can help difficult fits but does not guarantee a global optimum.
At a curvature join the Hessian assessment can be unavailable, which is
reported explicitly. The package does not silently replace a rejected
CV-selected candidate with another penalty.

## Scope and reproducibility

Inputs must be finite dense numeric matrices and a numeric response.
Missing values, constant training columns, and zero entry scores produce
errors. The package does not silently remove observations or predictors.
Formula interfaces, observation weights, sparse-matrix inputs and
non-Gaussian families are not implemented in this version.

Save the fold labels, input column order, package version and solver controls
with your analysis. When `foldid` is supplied, fitting does not consume random
numbers; otherwise folds are sampled from the caller's R random-number state.
No optional early stopping or one-standard-error selection rule is applied.

```{r version}
packageVersion("lambertReg")
```

The development repository is
[byuzbasi/lambertReg](https://github.com/byuzbasi/lambertReg). It is private
and requires access authorization; this package has not been released on
CRAN. Installation instructions are in the README. For reference details,
use `citation("lambertReg")`. The package is distributed under GPL-3.
