---
title: "Example gallery"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 10,
  fig.height = 6, out.width = "100%", dpi = 140)
library(lambertReg)
source(system.file("examples", "showcase.R", package = "lambertReg"))
demo <- lambert_showcase(tempfile("lambert-demo-"))
fit <- demo$fit
```

# One example, three views

The plots below use the same synthetic sample as the introductory guide:
100 observations, eight independent Gaussian predictors, three nonzero
coefficients, and seed 42. The first 80 observations form the training sample;
the last 20 are held out. Five-fold CV selects the penalty from its default
60-fraction path. These plots illustrate the package interface and are not
simulation summaries or real-data results from the manuscript.

## Choose a penalty

```{r cv, fig.alt="Cross-validation MSE with error bars, variable counts, and the selected penalty."}
plot(fit)
```

The orange marker identifies the selected penalty. Upper labels count selected
variables in the full-training models, excluding the intercept. Bars show one
fold-based standard error; overlapping training samples make them descriptive
rather than independent-sample confidence intervals.

## Follow the coefficients

```{r coefficients, fig.alt="Original-scale coefficient paths, including a marker at the selected penalty."}
plot(fit, type = "coefficients", labels = TRUE)
```

Each curve follows one original-scale coefficient. The intercept is excluded.
The dashed line marks the CV-selected position. Changing the displayed path
axis or plot style uses the stored fit and does not run another model.

## Prepare a compact figure

```{r paper-cv, fig.alt="Compact publication-style CV plot of the same stored fit."}
plot(fit, style = "paper")
```

The compact style keeps the CV curve, error bars, and variable counts while
removing the detailed title and selected-fit summary.

## Inspect the example's output

```{r summary}
knitr::kable(demo$summary, digits = 4, row.names = FALSE,
  col.names = c("Selected lambda", "Selected variables", "CV MSE", "Test RMSE", "Test MAE"))
```

The held-out errors describe only this small demonstration. They do not compare
Lambert with other estimators or establish a performance advantage.

The [data and reproduction guide](reproduction.html) gives the input dictionary
and the command that saves these plots, the folds, and the fit. The complete
[worked introduction](introduction.html) explains the modeling steps.
